Recognizing Strong Gravitational Lenses

### 2015-03-15 doc/Course_Paper/paper.tex
more updates

\usepackage{cvpr}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

% Include other packages here, before hyperref.

@@

%%%%%%%%% ABSTRACT
\begin{abstract}
    The detection of a large, representative set of strong gravitational lenses
    could greatly aid in our understanding of cosmology. Unfortunately they are
    quite rare, and the best techniques now revolve around squads of scientists
    manually scanning through images. This is presently borderline
    unsustainable and will be laughably inefficient with the advent of the
    Large Synoptic Survey Telescope. Here we examine the effectiveness of
    convolution neural networks and transfer learning for automated detection
    algorithms of strong gravitational lenses. We use images from the
    \textsc{Space Warps} project, a citizen science initiative to examine tens
    of thousands of fields of galaxies for the presence of strong gravitational
    lenses. 
    % TODO: Insert results!
    We find that using a convolution neural network trained on Galaxy
    morphologies as a feature extractor performs admirably but markedly worse
    than the citizen scientists.
    Scripts used in the analysis of this paper are freely available at
    \url{https://github.com/cpadavis/strongcnn}. The images are currently only
    available to those who contact the author, but will be available to the
    public in the near future.
\end{abstract}

@@

A consequence of Einstein's Theory of General Relativity
is that mass bends the path of light.\cite{Einstein:2007aa} Most of the time the deflections are very
small; the original 'gravitational lens' that tested the veracity of Einstein's
theory in the years after the first world war was the sun, which deflected
light from stars behind it only a few seconds of arc. However, when light
passes through a particularly deep gravitational potential (say, near the
center of the dark matter halo of a galaxy cluster), the deflections can be
particularly large, resulting in brilliant arcs and multiple images. These
strong deflections due to light passing through a deep gravitational potential
are termed strong gravitational lenses. The gravitational lensing signal can
heuristically be thought of as a trade off between a couple factors: larger
gradients in the gravitational potential create larger distortions (and larger
gradients in the gravitational potential tend to reside near to the center of
the foreground galaxy or galaxy cluster); the more separated the foreground and
background objects are, the bigger the proportional distortion on the
background object by the foreground (but the farther away the background object
is, the smaller it appears\footnote{Note that this is only true in the ``low
  redshift'' universe: when objects are farther away than a cosmological
  redshift of $z \approx 2.5$ or about 2.5 Gyr after the birth of the Universe,
they will actually \textit{grow} in angular extent.}).\cite{Kochanek:2006oz}

The very existence of these potentials acts as a verification of the Theory of
General Relativity, but they can also be used for much more. Strong
gravitational lenses are one of the few ways to directly probe the distribution
of dark matter, a particle (or possibly family of particles) that does not emit
electromagnetic radiation but does have mass and hence interacts
gravitationally with normal baryonic matter.\cite{Refsdal:1964aa} This allows us to tally the mass
of the largest gravitationally-bound structures in the universe, galaxy
clusters, which can give us insight into the formation history of these massive
objects. In this way, strong gravitational lenses can then tell us something
about the expansion history of the universe, by setting limits on how massive
the most massive objects in the universe can be. The properties of the bent
light itself can also say much about that expansion
history.\cite{Refsdal:1964ab, Kundic:1997aa, Barnacka:2014aa} When an object is
strongly-lensed into multiple images, each image travels a different span of
space and time. When an object does not vary much with time, these different
path lengths have no practical import. However, if the object varies
appreciably quickly (say it is a distant supermassive black hole at the center
of a galaxy whose accretion disk emits high-energy radiation at varying rates)
then these different path lengths can be used to pin down the rate of expansion
of the universe.

Unfortunately, for how useful strong gravitational lenses are, they are also
extremely rare. A next generation optical survey like the Large Synoptic Survey
Telescope can expect to find only ten thousand lenses in the whole sky, while
it will find ten billion galaxies. Currently in astronomy there are only order
hundreds of strong gravitational lenses known, mostly discovered by 'eyeball
squads' of graduate students. The small number means that target criteria must
be somewhat broad in order to maintain a relatively high completeness. Using
reasonable target criteria to find strong lenses such as looking only at
massive galaxies still means that nearly ten million objects will need to be
inspected in the next generation in order to find those ten thousand lenses. A
team of ten graduate students could expect to spend about 14 years looking at
these objects. Computer algorithms are not much better: most current machine
learning algorithms are woefully-underpowered for this task, and generally have
poor completeness or poor purity -- and often poor both. Additionally, some
algorithms are better at finding some types of lenses than others; some perform
well on the brilliant arcs, but poorly on the multiply-imaged objects, or vice
versa. For example, \cite{Kubo:2008aa} attempt to fit arc-like features in
images in order to find strong gravitational lenses, but this means that
multiply-imaged quasars are completely ignored. \cite{Agnello:2015aa} and
\cite{Chan:2014aa} in contrast develop an algorithm for finding gravitationally
lensed quasars based on catalog-level colors and magnitudes, precluding their
algorithm finding strong gravitational lens arcs. New algorithms need to be
developed to find more strong gravitational lenses, and more strong lenses need
to be found to power these algorithms. These algorithms need to not only
identify potential lenses accurately, but be able to make strong statements
about their contamination rates, as spectroscopic follow-up can be an expensive
endeavor. \cite{Momcheva:2015aa} performed spectroscopic follow-up on 9768
galaxies, finding 28 new strong gravitational lens systems, but taking 40
nights of telescope time on expensive telescopes.

@@

In this project we will use images collected by the Canada-France-Hawaii
Telescope Legacy Survey to analyze how Convolution Neural Networks can improve
automated detection of strong lens systems. We will also assess the performance
of citizen scientists by comparing our results to them. From other graduate
work (but not coursework), we have the locations and categories of around one
hundred and twenty known strong lenses, three thousand large fields verified to
contain no strong lenses, six thousand simulated strong lenses, and several
thousand classifications by citizen-scientists of other potential strong lens
systems.  These will form the core of our training and testing datasets; our
metric will be how well a CNN correctly identifies known and simulated lenses
and non-lens systems.

@@

From approximately 12000 fields of $440\times440\times3$ fields, we have
constructed approximately 30000 cutouts sized $96\times96\times3$. These cutouts
are selected based on where citizen scientists clicked, on the theory that both
'correct' and 'incorrect' selections provide useful information about the
characteristics of gravitational lenses. In general, we have access to two
broad classes of images: 'training' and 'test' images. The 'training' images
include fields that were verified in advance to not contain any lenses as well
as simulated lensed galaxies, quasars, and clusters. Many of the simulated
objects are over-exaggerated and extremely obvious, but we also have access to
a second 'refinement stage' of the project, where much harder simulations were
given to users. The 'test' images are the fields that citizen scientists
viewed, assessing whether a lens was in the field or not. In these 'test'
images are 120 known strong gravitational lens systems, which are also included
in this set. (The project confirms roughly half of these known lenses for
reasonable definitions of completeness and purity.) For all the images we also
have an associated probability that the project would evaluate that system as
containing a lens.

@@

The resultant Receiver Operating Curves can be observed in
Figure~\ref{fig:roc_tl}. In general we find that support vector machines
perform the best as a classifier on the feature vectors, but that the feature
vectors perform more poorly than the \textsc{Space Warps} users. We must caveat
though that \textsc{Space Warps} does not create a validation dataset against
which to test the performance of the system. Even if we don't perform quite as
well as the citizen scientists, we consider this a
promising baseline for future performance by transfer learning: we have not
even begun to consider potential performance gains by retraining the
convolution neural network on the \textsc{Space Warps} data.

@@

% Do we need this section?

@@

% Discuss how well things went. Or didn't go.
The need for new automated detection algorithms for finding strong
gravitational lenses will only become more pressing in the next decade, as it
becomes infeasible for scientists to scan images by eye for such systems. Using
the \textsc{Space Warps} dataset, we have examined how convolution neural
networks trained both on this particular dataset and on other datasets can
perform at the detection task.
% TODO: brief blurb about our CNN
We find that features extracted from a convolution neural network trained on
the classification of galaxy morphology (with a linear support vector
classifier on top for converting the feature vector to a binary ``lens'' and
``not-lens'') performs admirably but markedly worse than the citizen scientists
trained on the dataset. Further work examining improvements by retraining the
whole neural network could lead to a generic classification machine that takes
images from \textit{any} galaxy survey and states whether the image contains a
strong lens or not. Additionally there is much potential in both direct
convolution neural networks and transfer learning from other networks in
linking the classification outputs of the networks with the probability
estimates of the \textsc{Space Warps} system, which also examined several
thousands more ``unknown'' systems and could lead to more gravitational lenses
being identified.

%-------------------------------------------------------------------------
\section{Acknowledgements}
We received access to the dataset through Phil Marshall
\footnote{\url{pjm@slac.stanford.edu}}, who also graciously explained to us how
\textsc{Space Warps}
currently works. The convolutional neural network upon which the transfer
learning is based was kindly provided to us by Ryan Keisler
\footnote{\url{rkeisler@stanford.edu}}.
All errors in interpretation or otherwise are entirely our own.

>>>>

\usepackage{cvpr}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\input{macros.tex}
% Include other packages here, before hyperref.

@@

%%%%%%%%% ABSTRACT
\begin{abstract}
    The detection of a large, representative set of strong gravitational lenses
    could greatly aid in our understanding of cosmology. Unfortunately they are
    quite rare, and the best techniques now revolve around squads of scientists
    manually scanning through images. This is presently borderline
    unsustainable and will be laughably inefficient with the advent of the
    Large Synoptic Survey Telescope. Here we examine the effectiveness of
    convolution neural networks and transfer learning for automated detection
    algorithms of strong gravitational lenses. We use images from the
    \textsc{Space Warps} project, a citizen science initiative to examine tens
    of thousands of fields of galaxies for the presence of strong gravitational
    lenses. 
    % TODO: Insert results!
    We find that using a convolution neural network trained on Galaxy
    morphologies as a feature extractor performs admirably but markedly worse
    than the citizen-scientists.
    Scripts used in the analysis of this paper are freely available at
    \url{https://github.com/cpadavis/strongcnn}. The images are currently only
    available to those who contact the author, but will be available to the
    public in the near future.
\end{abstract}

@@

A consequence of Einstein's Theory of General Relativity is that mass bends the
path of light. \cite{Einstein:2007aa} Most of the time the deflections are very
small; the original 'gravitational lens' that tested the veracity of Einstein's
theory in the years after the first world war was the sun, which deflected
light from stars behind it only a few seconds of arc. However, when light
passes through a particularly deep gravitational potential (say, near the
center of the dark matter halo of a galaxy cluster), the deflections can be
particularly large, resulting in brilliant arcs and multiple images. These
strong deflections due to light passing through a deep gravitational potential
are termed strong gravitational lenses. The gravitational lensing signal can
heuristically be thought of as a trade off between a couple factors: larger
gradients in the gravitational potential create larger distortions (and larger
gradients in the gravitational potential tend to reside near to the center of
the foreground galaxy or galaxy cluster); the more separated the foreground and
background objects are, the bigger the proportional distortion on the
background object by the foreground (but the farther away the background object
is, the smaller it appears\footnote{Note that this is only true in the ``low
  redshift'' universe: when objects are farther away than a cosmological
redshift of $z \approx 2.5$ or about 2.5 Gyr after the birth of the Universe,
they will actually \textit{grow} in angular extent.}). \cite{Kochanek:2006oz}

The very existence of these potentials acts as a verification of the Theory of
General Relativity, but they can also be used for much more. Strong
gravitational lenses are one of the few ways to directly probe the distribution
of dark matter, a particle (or possibly family of particles) that does not emit
electromagnetic radiation but does have mass and hence interacts
gravitationally with normal baryonic matter. \cite{Refsdal:1964aa} This allows
us to tally the mass of the largest gravitationally-bound structures in the
universe, galaxy clusters, which can give us insight into the formation history
of these massive objects. \cite{Oguri:2010aa} They can also find extremely
distant objects. \cite{Hezaveh:2014aa} In this way, strong gravitational lenses
can then tell us something about the expansion history of the universe, by
setting limits on how massive the most massive objects in the universe can be.
The properties of the bent light itself can also say much about that expansion
history. \cite{Refsdal:1964ab, Kundic:1997aa, Gonzalez:2012aa, Barnacka:2014aa,
Hojjati:2014aa, Suyu:2014aa} When an object is strongly-lensed into multiple
images, each image travels a different span of space and time. When an object
does not vary much with time, these different path lengths have no practical
import. However, if the object varies appreciably quickly (say it is a distant
quasar, a supermassive black hole at the center of a galaxy whose accretion
disk emits high-energy radiation at varying rates) then these different path
lengths can be used to pin down the rate of expansion of the universe. Finally,
strong gravitational lenses also have the potential to rule out or validate
alternatives to General Relativity. \cite{Troxel:2014aa}

Unfortunately, for how useful strong gravitational lenses are, they are also
extremely rare. A next generation optical survey like the Large Synoptic Survey
Telescope or the Euclid space telescope can expect to find only ten thousand
lenses in the whole sky, while it will find ten billion galaxies.
\cite{Serjeant:2014aa} Currently in astronomy there are only order hundreds of
strong gravitational lenses known, mostly discovered by 'eyeball squads' of
graduate students. The small number means that target criteria must be somewhat
broad in order to maintain a relatively high completeness. Using reasonable
target criteria to find strong lenses such as looking only at massive galaxies
still means that nearly ten million objects will need to be inspected in the
next generation in order to find those ten thousand lenses. A team of ten
graduate students could expect to spend about 14 years looking at these
objects. Computer algorithms are not much better: most current machine learning
algorithms are woefully-underpowered for this task, and generally have poor
completeness or poor purity -- and often poor both. Additionally, some
algorithms are better at finding some types of lenses than others; some perform
well on the brilliant arcs, but poorly on the multiply-imaged objects, or vice
versa. For example, \cite{Kubo:2008aa} attempt to fit arc-like features in
images in order to find strong gravitational lenses, but this means that
multiply-imaged quasars are completely ignored. \cite{Agnello:2015aa} and
\cite{Chan:2014aa} in contrast develop an algorithm for finding gravitationally
lensed quasars based on catalog-level colors and magnitudes, precluding their
algorithm finding strong gravitational lens arcs. New algorithms need to be
developed to find more strong gravitational lenses, and more strong lenses need
to be found to power these algorithms. These algorithms need to not only
identify potential lenses accurately, but be able to make strong statements
about their contamination rates, as spectroscopic follow-up can be an expensive
endeavor. \cite{Momcheva:2015aa} performed spectroscopic follow-up on 9768
galaxies, finding 28 new strong gravitational lens systems, but taking 40
nights of telescope time on expensive telescopes.

@@

In this project we will use images collected by the Canada-France-Hawaii
Telescope Legacy Survey to analyze how Convolution Neural Networks can improve
automated detection of strong lens systems. We will also assess the performance
of citizen-scientists by comparing our results to them. From other graduate
work (but not coursework), we have the locations and categories of around one
hundred and twenty known strong lenses, three thousand large fields verified to
contain no strong lenses, six thousand simulated strong lenses, and several
thousand classifications by citizen-scientists of other potential strong lens
systems.  These will form the core of our training and testing datasets; our
metric will be how well a CNN correctly identifies known and simulated lenses
and non-lens systems.

@@

From approximately 12000 fields of $440\times440\times3$ fields, we have
constructed approximately 30000 cutouts sized $96\times96\times3$. These cutouts
are selected based on where citizen-scientists clicked, on the theory that both
'correct' and 'incorrect' selections provide useful information about the
characteristics of gravitational lenses. In general, we have access to two
broad classes of images: 'training' and 'test' images. The 'training' images
include fields that were verified in advance to not contain any lenses as well
as simulated lensed galaxies, quasars, and clusters. Many of the simulated
objects are over-exaggerated and extremely obvious, but we also have access to
a second 'refinement stage' of the project, where much harder simulations were
given to users. The 'test' images are the fields that citizen-scientists
viewed, assessing whether a lens was in the field or not. In these 'test'
images are 120 known strong gravitational lens systems, which are also included
in this set. (The project confirms roughly half of these known lenses for
reasonable definitions of completeness and purity.) For all the images we also
have an associated probability that the project would evaluate that system as
containing a lens.

@@

The resultant Receiver Operating Curves can be observed in
Figure~\ref{fig:roc_tl}. In general we find that support vector machines
perform the best as a classifier on the feature vectors, but that the feature
vectors perform more poorly than the \textsc{Space Warps} users. We must caveat
though that \textsc{Space Warps} does not create a validation dataset against
which to test the performance of the system. Even if we don't perform quite as
well as the citizen-scientists, we consider this a
promising baseline for future performance by transfer learning: we have not
even begun to consider potential performance gains by retraining the
convolution neural network on the \textsc{Space Warps} data.

@@

We briefly attempt to answer the questions we posed at the beginning of this
study:

% answer the questions
\subsection*{Do we have enough data to reasonably train and test a CNN?}

We do, and we have easy ways to augment the data through rotations, flips, and
(in the future) other transformations that leave the identification of a strong
gravitational lens invariant.

\subsection*{How do citizen-scientists do compared with this automated system?}

Currently the citizen-scientists outperform our automated systems.

\subsection*{Can we use the results of citizen-scientists to train the CNN?}

In the sense that the citizen-scientists are used in creating the cutouts for
our system, we find their results quite useful. We were unable to use them for
much more than that, however. Future work could involve calibrating our
classifications against the probabilities from the \textsc{Space Warps}
citizen-scientist classification system.

\subsection*{How well does using features extracted from a convolution neural
network trained on galaxy morphology perform when determining the presence of
strong gravitational lenses?}

We appear to do admirably well in that our receiver operating curves obtain
remarkably high true positive rates for modestly low false positive rates.
However, the extracted features do worse than the citizen-scientists from
\textsc{Space Warps}.

@@

% Discuss how well things went. Or didn't go.
The need for new automated detection algorithms for finding strong
gravitational lenses will only become more pressing in the next decade, as it
becomes infeasible for scientists to scan images by eye for such systems. Using
the \textsc{Space Warps} dataset, we have examined how convolution neural
networks trained both on this particular dataset and on other datasets can
perform at the detection task.
% TODO: brief blurb about our CNN
We find that features extracted from a convolution neural network trained on
the classification of galaxy morphology (with a linear support vector
classifier on top for converting the feature vector to a binary ``lens'' and
``not-lens'') performs admirably but markedly worse than the citizen-scientists
trained on the dataset. Further work examining improvements by retraining the
whole neural network could lead to a generic classification machine that takes
images from \textit{any} galaxy survey and states whether the image contains a
strong lens or not. Additionally there is much potential in both direct
convolution neural networks and transfer learning from other networks in
linking the classification outputs of the networks with the probability
estimates of the \textsc{Space Warps} system, which also examined several
thousands more ``unknown'' systems and could lead to more gravitational lenses
being identified.

%-------------------------------------------------------------------------
\subsection*{Acknowledgements}
We received access to the dataset through Phil Marshall
\footnote{\url{pjm@slac.stanford.edu}}, who also graciously explained to us how
\textsc{Space Warps}
currently works. The convolutional neural network upon which the transfer
learning is based was kindly provided to us by Ryan Keisler
\footnote{\url{rkeisler@stanford.edu}}.
All errors in interpretation or otherwise are entirely our own.
